A simulation model of DNA methylation data

### 2015-03-30 Rmarkdown/methsim.Rmd
Clarify how LOR is sampled for rare/impossible region-IPD combinations. Add data for claims made in this commit.

These same distributions for all the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ samples are shown in Figures \ref{fig:EPISCOPE_beta_by_pm_boxplots}, \ref{fig:EPISCOPE_lor_by_pm_quantile_plots}, \ref{fig:Lister_beta_by_pm_boxplots}, \ref{fig:Lister_lor_by_pm_quantile_plots},
\ref{fig:Seisenberger_beta_by_pm_boxplots}, \ref{fig:Seisenberger_lor_by_pm_quantile_plots},
\ref{fig:Ziller_merged_beta_by_pm_boxplots} and \ref{fig:Ziller_merged_lor_by_pm_quantile_plots} (__TODO: Are these other examples necessary?__).

@@

The second step requires further explanation. If there are an insufficient number of observations with the given $IPD$-region combination, then, by default, `methsim` samples $\Psi_{(i, i + 1)}$ from a Gaussian distribution[^loess_params]. By default, if there are less than __TOOD: how many?__ $LOR$ with the given combination of region-type and IPD then the Gaussian-approximation is used. The user may specify a different method for dealing with these rarer $IPD$-region combinations and, indeed, different methods are explored in the __RESULTS SECTION__.

[^loess_params]: The mean, $\mu$, and standard deviation, $\sigma$, of this Gaussian distribution depend on both $IPD$ and $r_{i}$. A loess fit of $mean(LOR)$ (resp. $sd(LOR)$) as a function of $IPD$ is performed, stratified by region type. The values of $\mu$ (resp. $\sigma$) are then obtained by extrapolating the loess fit (__TODO: Update if changes made to `methsim` defaults__).

@@

With all the above limitations in mind, `methsim` still uses the simple strategy of sampling the $\mathbf{W}$ from $w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}$. At present I do not have a better alternative and, as we shall see, the current procedure still procedure still produced realistic simulated data. In the __RESULTS SECTION__ I also explore a simulation that simulates $n_{\bar{h}}$ independent realisations of $\mathbf{Z}$ where $n_{\tilde{h}} < n_{\bar{h}} \\l n_{h}$ and uniformly sets $\mathbf{W_{r_{i}, \cdot}} = (\frac{1}{n_{\bar{h}}, \ldots, \frac{1}{n_{\bar{h}})$.

>>>>

These same distributions for all the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ samples are shown in Figures \ref{fig:EPISCOPE_beta_by_pm_boxplots}, \ref{fig:EPISCOPE_lor_by_pm_quantile_plots}, \ref{fig:Lister_beta_by_pm_boxplots}, \ref{fig:Lister_lor_by_pm_quantile_plots},
\ref{fig:Seisenberger_beta_by_pm_boxplots}, \ref{fig:Seisenberger_lor_by_pm_quantile_plots},
\ref{fig:Ziller_merged_beta_by_pm_boxplots} and \ref{fig:Ziller_merged_lor_by_pm_quantile_plots}.

@@

The second step requires further explanation. If there are no observations with the required $IPD$-region combination, then, by definition, `methsim` is unable to sample from the appropriate empirical distribution[^min_n]. Instead, when this occurs, `methsim` samples from a user-specified parametric distribution. __RESULTS SECTION__ explores different choices of this distribution __AND SHOWS WHAT__. The default option is a zero-mean Gaussian distribution, chosen so that "on average" the methylation state at adjacent CpGs with large $IPD$s are independent of one another. The default standard deviation of this Gaussian distribution is $2$; this value, while fairly arbitrary, reflects the median standard deviation of the $LOR_{(i, i+1) | r_{i}, IPD(i, i+1)}$ distributions for each sample, which is remarkably constant across $r_{i}$ and samples[^ipd_stability] (__TODO: Format and reference table__).

[^min_n]: By default, `methsim` requires at least 100 observations in the  $LOR_{(i, i+1) | r_{i}, IPD(i, i+1)}$ to try to avoid sampling from sparse empirical distributions.

[^ipd_stability]: Figures \ref{fig:EPISCOPE_lor_by_pm_quantile_plots}, \ref{fig:Lister_lor_by_pm_quantile_plots}, \ref{fig:Seisenberger_lor_by_pm_quantile_plots} and \ref{fig:Ziller_merged_lor_by_pm_quantile_plots} show that the variability of these distributions is fairly constant across $IPD$ for a given $r_{i}$ and sample.

```{r, eval = TRUE, echo = FALSE}
knitr::kable(readRDS("../data/median_sd_LOR_distributions.rds"), align = "c", digits = 2)
```

@@

With all the above limitations in mind, `methsim` still uses the simple strategy of sampling the $\mathbf{W}$ from $w_{(i, \ldots, i + m - 1) | r_{(i, \ldots, i + m - 1)}}$. At present I do not have a better alternative and, as we shall see, the current procedure still procedure still produced realistic simulated data. In the __RESULTS SECTION__ I also explore a simulation that simulates $n_{\bar{h}}$ independent realisations of $\mathbf{Z}$ where $n_{\tilde{h}} < n_{\bar{h}} \\l n_{h}$ and uniformly sets $\mathbf{W_{r_{i}, \cdot}} = (\frac{1}{n_{\bar{h}}}, \ldots, \frac{1}{n_{\bar{h}}})$.
